---
title: "Unsupervised learning examples"
author: "C.A. Kapourani"
output: 
  html_document: 
    highlight: haddock
    theme: cerulean
    number_sections: true
    toc: yes
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, results = "hide")
```

# Generate data from Gaussian distribution
First we show the density of a Gaussian distribution, with mean = 1, and standard deviation = 1.
```{r}
# Density of Gaussian distribution
x <- seq(-1, 3, length = 1000)
hx <- dnorm(x, mean = 1, sd = 1)

plot(x, hx, type = "l", lty = 1, xlab="x", 
     ylab = "p(x)", col = "red3", lwd = 3)
```

Now we generate samples from this normal distribution and plot a histogram.
```{r}
x <- rnorm(10000, mean = 1, sd = 1)
mu <- sum(x) / length(x)
sigma <- sd(x)
hist(x)
```

# PCA as data compression algorithm
This example mostly follows this blog: https://rpubs.com/aaronsc32/image-compression-principal-component-analysis

First we load the jpeg image and since it is a coloured image, we store each channel in a different matrix. The cat image is represented as three 600x398 matrices as an array with each matrix corresponding to the RGB color value scheme.
```{r}
library(jpeg)
cat <- readJPEG('figures/cat.jpg')

# Extract RGB colour
r <- cat[,,1]
g <- cat[,,2]
b <- cat[,,3]
```

Next we run PCA (by calling the `prcomp` function) on each colour channel independently and finally colect everything in a list.
```{r}
# Perform PCA on each colour value matrix
cat.r.pca <- prcomp(r, center = FALSE)
cat.g.pca <- prcomp(g, center = FALSE)
cat.b.pca <- prcomp(b, center = FALSE)

# Collect everything in a list
rgb.pca <- list(cat.r.pca, cat.g.pca, cat.b.pca)
```

Reconstruct the images from the kth principal component.
```{r}
# How many components to use to reconstruct the original image
pcs <- c(1, 2, 4, 10, 15, 46, 89, 131, 345)

#for (i in seq.int(3, round(nrow(cat) - 10), length.out = 10)) {
for (i in pcs){
  pca.img <- sapply(rgb.pca, function(j) {
    compressed.img <- j$x[,1:i] %*% t(j$rotation[,1:i])
  }, simplify = 'array')
  writeJPEG(pca.img, paste('figures/cat_comp_', round(i,0), '_pcs.jpg', sep = ''))
}
```


Finally we can compute the proportion of variance explained. We just show the top 30 PCs for visualisation purposes.
```{r}
var_explained <- cat.b.pca$sdev / sum(cat.b.pca$sdev)
plot(seq(1:30), var_explained[1:30], type = "l")
points(seq(1:30), var_explained[1:30])
```

